---
title: "adapref_intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{adapop_intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adapref)
```

We will first define two effective population size functions (both exponential but distinct in this case), and a sampling intensity (uniform). 
Note: you need to have installed phylodyn, and ape. 

```{r}

set.seed(123)
n<-100

exp_traj1<-function(t){
  result=rep(0,length(t))
  result[t<=0.35]<-10
  result[t>0.35 & t<0.4]<-9.999987e+14*exp(-92.1034*t[t>0.35 & t<0.4]) #50 times smaller
  result[t>=0.4]<-0.1
  return(result)
}

exp_traj2<-function(t){
  result=rep(0,length(t))
  result[t<=0.1696721]<-10
  result[t>0.1696721 & t<0.2196721]<-61218647*exp(-92.1034*t[t>0.1696721 & t<0.2196721]) #50 times smaller
  result[t>=0.2196721]<-0.1
  return(result)
}

coeff=n/0.4
sampling_traj<-function(t){
  result=rep(0,length(t))
  result<-1
  return(result*coeff)}

```

We now sample the sampling times, coalescent times, and a corresponding genealogy. We repeat it twice to account for the two populations. 
Note that in the samp_times vector, we artificially make the first sampling time zero.
n_sampled is going to be a vector of 1, because the Poisson process samples one event a time

We then plot the simulated genealogies. 
```{r}
samp_times1<-c(0,sampsim_thin(max.n=n-2,traj=sampling_traj,xlim=c(0,3))[-n+1])
n_sampled1 <- c(2,rep(1,n-2)) # simulation like this it will be like this.
samp_times2<-c(0,sampsim_thin(max.n=n-2,traj=sampling_traj,xlim=c(0,3))[-n+1])
n_sampled2 <- c(2,rep(1,n-2)) # simulation like this it will be like this.

simulation1<-phylodyn::coalsim(samp_times=samp_times1-min(samp_times1),n=n_sampled1,traj=exp_traj1)
tree1<-phylodyn::sample_genealogy(simulation1)
tree1<-ape::read.tree(text=tree1$Newick)

simulation2<-phylodyn::coalsim(samp_times2-min(samp_times2),n_sampled2, traj = exp_traj2)
tree2<-phylodyn::sample_genealogy(simulation2)
tree2<-ape::read.tree(text=tree2$Newick)
coal_times1last <- max(simulation1$coal_times)
coal_times2last <- max(simulation2$coal_times)


plot(tree1,show.tip.label = FALSE)
plot(tree2,show.tip.label = FALSE)

```

We then estimates effective population sizes with the function BNPR_adapop. We first fit the nonparametric model for the dependence (parSel=F) and don't use preferential sampling (preferential=F)
```{r}
res<-BNPR_pop(tree1,tree2,samp_times1,samp_times2,lengthout=100,parSel = F,
                  preferential =F)
```


You may get some warnings, but these are mostly INLA related warnings. So, don't worry!

Now, we plot the outputs. With the function plot_BNPR_plus, we can plot all different posterior distribution just changing the name of the parameter we want to plot.
Note: plotting the population size of the second population size takes longer because you need to sample from the posterior

```{r,echo=T}
plot_BNPR_plus(res,heatmaps=FALSE,heatmap_labels = FALSE,ylim=c(0.01,1000),xlim=c(0.75,0),parameter = "eff1")
plot_BNPR_plus(res,heatmaps=FALSE,heatmap_labels = FALSE,ylim=c(0.01,1000),xlim=c(0.75,0),parameter = "eff2")
plot_BNPR_plus(res,heatmaps=FALSE,heatmap_labels = FALSE,ylim=c(0.001,1000),xlim=c(0.75,0),parameter = "gamma")

```


We can extend the above function to the parameteric model (see reference for more details) or preferential sampling. We try here preferential sampling and compare (visually) the estimates to the nonpreferential sampling scenario. 

```{r}
res_ps<-BNPR_pop(tree1,tree2,samp_times1,samp_times2,lengthout=100,parSel = F,
                  preferential =TRUE)

plot_BNPR_plus(res,heatmaps=FALSE,heatmap_labels = FALSE,ylim=c(0.01,1000),xlim=c(0.75,0),parameter = "eff1")
plot_BNPR_plus(res_ps,heatmaps=FALSE,heatmap_labels = FALSE,ylim=c(0.01,1000),xlim=c(0.75,0),parameter = "eff1")

```

When one uses preferential sampling, it is possible to obtain a plot of the posterior sampling intensity. 
